---
title: "Homework 7--Qingqing"
output:
  html_document:
    df_print: paged
---

```{r}
par(mfrow=c(2,4))

#First I play around with my data set a bit to see if I notice any correlations between chlorophyll a and other variables based I what I know about chlorophyll
plot(wq$`Chlorophyll a`)
plot(wq$`1% Light Depth`, wq$`Chlorophyll a`)
plot(wq$Depth, wq$`Chlorophyll a`)
#Since chlorophyll is a light capturing component of plant cells I would expect less of it to exist at greater depths where sunlight penentration decreases  It's usuless for plant cells to have more chlorophyll in low light conditions waste of resources and energy
plot(wq$`Manganese (Dissolved)`,wq$`Chlorophyll a`)
plot(wq$`Manganese (Total)`, wq$`Chlorophyll a`)
#Magnesium is a major building block for chlorophyll in general, so the more of that that is readily avaialable for plant cells to take in and mineralize, the more resources they have to create more chlorphyll if necessary
plot(wq$`Ammonia (Total)`, wq$`Chlorophyll a`) #High pH and ammonia has a negative effect on photosynthesis and can damage plant cells.  Whule plants can take in ammonia as a form of nitrogen, it needs to be materialized rapidly as ammonia in its current form is toxic to the plant cell.  I feel in an environment where pH is high and ammonia concentrations are high, than plant cells will not thrive as well and major die offs are a possibility leading to a decrease in Chlrophyll
plot(wq$pH, wq$`Chlorophyll a`)
plot(wq$Oxygen, wq$`Chlorophyll a`)
#Since within the chlorophyll during light dependent reactions, water gets split into hydrogen and O2 so I feel the greater concentration of plant cells and chlorophyll the more oxygen that should be released

#While some of these plots are showing up blank for me during my pre-tests, I definitely can notice signs of relationships and correlations between some of the variables like depth and magnesium vs. chlorophyll concentrations.  These are variables I will look into using for my final predictive model

```


```{r}
library(tidyverse)
library(lubridate)
```

Question 1. For the Water Years 2005-2012, develop a predictive model of the mean monthly Chl-a concentration in the California Bay Delta using other mean monthly water quality variables.

Step 1: get my desired dataset

```{r}
BayDeltaWQ <- read.csv("BayDeltaWQ.csv")
BayDeltaWQ
names(BayDeltaWQ)
#Let's first subset the data Water Years 2005-2012 (October 1, 2004 and ended on September 30, 2012)
BayDeltaWQ$Date <- ymd(BayDeltaWQ$SampleDate)
wtr_yr <- function(x) {
  offset = ifelse(month(x) >= 10, 1, 0)
  adj.year = year(x) + offset
  adj.year
}

BayDeltaWQ$Water_Year <- wtr_yr(BayDeltaWQ$Date)

WQ <- subset(BayDeltaWQ, Water_Year >= 2005 & Water_Year <= 2012 )
# Double check if we get the desired dataset
range(WQ$Water_Year)
range(as.character(WQ$SampleDate))
range(WQ$Date)
#Add two columns: year, month
WQ$year <- year(WQ$Date)
WQ$month <- month(WQ$Date)
#I would like to delete the columns with all NAs and then select the water quality variables that I can calculate the mean monthly value since they do not make sense in this analysis.
WQ2 <- WQ[colSums(!is.na(WQ)) > 0]
names(WQ2)
WQ2 <- WQ2[,-c(1:3,7,16,33:34)]
WQ2$Depth <- as.numeric(as.character(WQ2$Depth))
test <- WQ2[complete.cases(WQ2),]
#There is no observation if I remove the observations with any NAs. Let's see how many NAs each variable has
NA_Num <- data.frame(sapply(WQ2, function(x) sum(is.na(x))))

#Aggregate to mean monthly data
WQ3<- WQ2 %>%
    group_by(year,month)%>%
    summarise_all(funs(mean(., na.rm=TRUE)))
NA_Num <- data.frame(sapply(WQ3, function(x) sum(is.na(x))))
#subset to delete the variables that contains NA
WQ3 <- subset(WQ3, select=colMeans(is.na(WQ3)) == 0)
# OK! WQ3 is my final dataset! Let's take a look
range(WQ3$year)
names(WQ3)
save(WQ3,file="WQ3.rda")

```

Step 2: Identify variables that are highly correlated with the others

 What is Chl-a concentration? BayDeltaWQ$Chlorophyll.a
 https://www.epa.gov/national-aquatic-resource-surveys/indicators-chlorophyll
```{r}
#Delete two columns: year and month. And move the Chlorophyll.a to be the first column
Water <- WQ3[,-(1:2)]%>%
  select(Chlorophyll.a, everything())
a <- cor(Water)
unique(Water$Depth)
#Delete the variable "Depth" since the standard deviation is zero (all the values are 3)
Water <- Water[,-2]
names(Water)
```

Save for the future! (Some of my thought, like data transformation that I didn't do in this analysis...)

Take a look at the data WQ3$Chlorophyll.a. The data is normally distributed after log transformation. 
 plot(density(WQ3$Chlorophyll.a))
 plot(density(log(WQ3$Chlorophyll.a)))
 qqnorm(WQ3$Chlorophyll.a);qqline(WQ3$Chlorophyll.a)
 qqnorm(log(WQ3$Chlorophyll.a));qqline(log(WQ3$Chlorophyll.a))
 
P <- apply(Water,2, function(col)cor(col, Water$Chlorophyll.a,method = "pearson"))

K <- apply(Water,2, function(col)cor(col, Water$Chlorophyll.a,method = "kendall"))

S <- apply(Water,2, function(col)cor(col, Water$Chlorophyll.a,method = "spearman"))

P_cor <- cor(Water,method = "pearson")
K_cor <- cor(Water,method = "kendall")
S_cor <- cor(Water,method = "spearman")


```{r}
# A correlation between -1.0 to -0.5 or 0.5 to 1.0 means that variables are highly correlated.
Fun_corr <- function(x, y) {
  z <- upper.tri(x)
  data.frame(
    row = rownames(x)[row(x)[z]],
    column = rownames(x)[col(x)[z]],
    cor = x[z],
    p = y[z]
    )
}
library(Hmisc)
cor_water <-rcorr(as.matrix(Water),type="spearman") #rcorr type should be one of “pearson”, “spearman”. I will go with the nonparamertic method "spearman" and keep with further analysis for consistency
test <- Fun_corr(cor_water$r, cor_water$P) 

High_cor <- subset(test, abs(cor) >= 0.5)
High_cor  #High_cor shows the variables that are highly correlated to others
#Chlorophyll.a is highly correlated with Temperature (cor: 0.6003798,P: 1.009286e-10),Ammonia..Dissolved.(cor:-0.5393164,P:1.433183e-08) and Pheophytin.a (cor: 0.7016413, P:1.776357e-15). 
cor(Water$Temperature,Water$Pheophytin.a)
#0.3554389 
cor(Water$Temperature,Water$Ammonia..Dissolved.)
#-0.6195209
cor(Water$Pheophytin.a,Water$Ammonia..Dissolved.)
#-0.1262
#There are correlations among Temperature, Pheophytin.a. and Ammonia..Dissolved.Let's do a partial correlation with 4 variables Chlorophyll.a, Temperature, Pheophytin.a. and Ammonia..Dissolved.
model_data <- dplyr::select(Water, Chlorophyll.a, Temperature, Pheophytin.a, Ammonia..Dissolved.)
save(model_data,file="model_data.rda")
require(ppcor)
mypcor <- pcor(model_data, method = "spearman")
mycor <- cor(model_data, method = "spearman")
require(corrplot)
corrplot(mycor, method = "number")
#Conclusion: Much of the apparent correlation between Chlorophyll.a and Temperature is explained by the mutual negative correlation with Ammonia..Dissolved..Pheophytin.a and Ammonia..Dissolved. should be the best predictors of Chlorophyll.a.
```

Step 3: develop a predictive model of the mean monthly Chl-a concentration

I developed the predictive models in two ways. 

(1) Develop a model with variables that were highly correlated with Chlorophyll.a. Based on previous correlation analysis, Pheophytin.a and Ammonia..Dissolved. should be the best predictors of Chlorophyll.a. 
My null model (lm.null)is that every value is predicted by the mean.
The other three models are:lm.C.P which only use Pheophytin.a to predict Chlorophyll.a;
lm.C.A which only use Ammonia..Dissolved. to predict Chlorophyll.a;
lm.C.A.P which use both Ammonia..Dissolved. and Pheophytin.a to predict Chlorophyll.a.

Results 1: Since the modelthat use both Ammonia..Dissolved. and Pheophytin.a to predict Chlorophyll.a. has the lowest AIC and BIC value, I would choose this model! 


(2) Do a step model with all the variables that are available to develop the model.

Result 2: I selected the model lm(formula = Chlorophyll.a ~ Conductance..EC. + Oxygen + Temperature + Ammonia..Dissolved. + Kjeldahl.Nitrogen..Total. + Organic.Nitrogen..Dissolved. + 
    Pheophytin.a + Solids..Total.Dissolved., data = Water) that has the lowest AIC value.

Develop model with highly correlated variables
```{r}
lm.null <- lm(Chlorophyll.a ~ 1, data = model_data); summary(lm.null)
lm.C.P <- lm(Chlorophyll.a ~ Pheophytin.a, data = model_data); summary(lm.C.P)
lm.C.A <- lm(Chlorophyll.a ~ Ammonia..Dissolved., data = model_data); summary(lm.C.A )
lm.C.A.P <- lm(Chlorophyll.a ~ Ammonia..Dissolved. + Pheophytin.a, data = model_data); summary(lm.C.A.P)
require(tidyverse)
require(broom)
lms <- list(null=lm.null, C.P = lm.C.P, C.A = lm.C.A, C.A.P = lm.C.A.P)
lms.stats <- mapply(glance, lms)
colnames(lms.stats) <- names(lms)
lms.stats
lms.coeffs <- lapply(lms, tidy)
lms.coeffs
lms.coeffs.notalist <- Map(cbind, lms.coeffs, model = names(lms))
lms.coeffs.notalist <- bind_rows(lms.coeffs.notalist)
lms.coeffs.notalist
unlist(lms.stats["adj.r.squared",])
aic <- as.data.frame(unlist(lms.stats["AIC",]))
colnames(aic) <- "AIC"
aic
bic <- as.data.frame(unlist(lms.stats["BIC",]))
colnames(bic) <- "BIC"
bic
aic.bic <- cbind(aic,bic)
a.b <- aic.bic %>% 
  rownames_to_column() %>% 
  gather(metric, value, -rowname) %>% 
  ggplot(aes(rowname, value, fill = metric)) +
  geom_bar(position = "dodge", stat = "identity")
a.b
# The modelthat use both Ammonia..Dissolved. and Pheophytin.a to predict Chlorophyll.a. has the lowest AIC and BIC value
```

Step model
```{r}
lms <- step(lm(Chlorophyll.a ~ ., data = Water))
final.mod <- stepAIC(lms)
```

Step 4: Final model selection
My final model: lm(formula = Chlorophyll.a ~ Conductance..EC. + Oxygen + Temperature + Ammonia..Dissolved. + Kjeldahl.Nitrogen..Total. + Organic.Nitrogen..Dissolved. + 
    Pheophytin.a + Solids..Total.Dissolved., data = Water)
Diagnostic Plots analysis:
(1)Residual plots indicate that both models meet the regression assumptions very well.
(2)From the Q-Q plots we can see residuals are Normally distributed. Observations numbered as 57 and 45 are extreme values. 
(3)From Scale-Location plots, final.mod model perfoms better because it has a horizontal line with equally (randomly) spread points.
(4)Residuals vs Leverage plots help us to find influential cases.The plot identified the influential observation as #15. If I exclude the 15th case from the analysis, the slope coefficient and R2 would change.
From the analysis above, I will go with the final.mod model. Let's take a look at the comparasion of predicted values and the actual values.   

Reference:https://www.statmethods.net/stats/rdiagnostics.html
          https://onlinecourses.science.psu.edu/stat501/node/36
          http://data.library.virginia.edu/diagnostic-plots/
          
```{r}
chl_stats <- t(tibble(rmse = rmse(wq2com$`Chlorophyll a`,wq2com$`Chlorophyll a`.fromC), 
                     mae = mae(wq2com$`Chlorophyll a`,wq2com$`Chlorophyll a`.fromC), 
                     pbias = pbias(wq2com$`Chlorophyll a`,wq2com$`Chlorophyll a`.fromC), 
                     cor = cor(wq2com$`Chlorophyll a`,wq2com$`Chlorophyll a`.fromC)))
stats <- cbind(load.stats, conc.stats)
```

```{r}
#First, take a look at the results.
summary(final.mod)
summary(lm.C.A.P)
#final.mod has a higher value of R-square (0.593) compared with lm.C.A.P (0.4728).

layout(matrix(c(1,2,3,4),2,2))
plot(final.mod)
plot(lm.C.A.P)

Water$predicted <- predict(final.mod)   # Save the predicted values
Water$residuals <- residuals(final.mod)
plot(Water$Chlorophyll.a, Water$predicted,
      xlab="actual", ylab="predicted")
 abline(a=0,b=1)
save(Water,file="Water.rda")
```


Step 5: Quantify how much variance in mean monthly Chl-a is explained by the variable selected.
Results: 59.3% variance in mean monthly Chl-a is explained by the variable selected. Pheophytin.a has the highest relative importance

Reference: https://stats.stackexchange.com/questions/79399/calculate-variance-explained-by-each-predictor-in-multiple-regression-using-r
```{r}
summary(final.mod)
#R-square is the percentage of the response variable variation that is explained by the variable selected. In this model R-square is 0.593 which means that 59.3% variance in mean monthly Chl-a is explained by the variable selected.

#Take a look at the relative importance of the variables.
library(relaimpo)
calc.relimp(final.mod,type=c("lmg","last","first","pratt"),
   rela=TRUE)
# Bootstrap Measures of Relative Importance (1000 samples) 
boot <- boot.relimp(final.mod, b = 1000, type = c("lmg", 
  "last", "first", "pratt"), rank = TRUE, 
  diff = TRUE, rela = TRUE)
booteval.relimp(boot) # print result
layout(matrix(c(1,2,3,4),2,2))
plot(booteval.relimp(boot,sort=TRUE)) # plot result
# Results indicate that Pheophytin.a has the highest relative importance.
```

Question 2: Model the mean monthly Chl-a concentration as a function of season and the water quality variable idenitifed as MOST important in explaining the variance of Chl-a from question 1

Step 1:create a new discrete value called "season", and assign a label to  data for "wet season" and "dry season". Model the mean monthly Chl-a concentration as a function of season and Pheophytin.a.

Answer: 
There is no clear defination of dry and wet season for Bay Delta. According to the monthly precipitation in Sacramento from https://www.usclimatedata.com/climate/california/united-states/3174, I will define precipiation equal or larger than 1 inch to be the wet season (Novermber to April) and lower than inch to be the dry season (May to October) in this analysis.
14.72%  variation in Chlorophyll.a is explained by season. 
35.58%  variation in Chlorophyll.a is explained by Pheophytin.a.
40.86%  variation in Chlorophyll.a is explained by both season and Pheophytin.a.
The combined model is better than individual models.
The mean of Chlorophyll.a in dry season is 4.58 while in wet season is 2.82.
The slopes of the linear regression after accounting for season are: wet season (-1.0952),Pheophytin.a(1.7).
The slope of the linear regression not considering season is Pheophytin.a(1.9083).
From the plots we can see residuals are normally distributed (but with several extreme events). 

Reference: https://feliperego.github.io/blog/2015/10/23/Interpreting-Model-Output-In-R
```{R}
WQ3$Season <- ifelse(WQ3$month %in% c(5:10), "DrySeason",
                        "WetSeason"  )
WaterSeason <- dplyr::select(WQ3, Chlorophyll.a, Pheophytin.a, Season)
WaterSeason$Season <- as.factor(WaterSeason$Season)
save(WaterSeason,file="WaterSeason.rda")

lm.C.S <- lm(Chlorophyll.a ~ Season, data = WaterSeason); summary(lm.C.S)
lm.C.P <- lm(Chlorophyll.a ~ Pheophytin.a, data = WaterSeason); summary(lm.C.P)
lm.C.S.P <- lm(Chlorophyll.a ~ Season + Pheophytin.a, data = WaterSeason); summary(lm.C.S.P)

DrySeason <- mean(WaterSeason[WaterSeason$Season %in% "DrySeason",]$Chlorophyll.a)
WetSeason <- mean(WaterSeason[WaterSeason$Season %in% "WetSeason",]$Chlorophyll.a)

plot(lm.C.S.P)
```

Step 2: Plot Chl-a vs your most important variable, with the observations colored by season. Add the paralell regression lines from the combined model in yellow, and add the univariate regression line in blue.
```{r}
predicted_df <- data.frame(Chl_pred = predict(lm.C.S.P, WaterSeason), Pheophytin.a = WaterSeason$Pheophytin.a)

ggplot(WaterSeason) +
      geom_point(mapping = aes(x = Pheophytin.a , y = Chlorophyll.a, color = Season)) +
  geom_smooth(method = "lm", mapping = aes(x = Pheophytin.a , y = Chlorophyll.a))+
geom_line(color='yellow',data = predicted_df, aes(x=Pheophytin.a , y= Chl_pred))
```
